Estimating the number of clusters using Cross Validation

### 2015-06-19 cvclust.tex
reorg k-means proof

Our proof relies on results from \citet{pollard1981strong}, who analyzed
$k$-means and showed that the within-cluster sum of squares converges to a
population quantity.  Specifically, for any finite set $A \subset
\mathbb{R}^d$ and probability measure $P$ supported on $\mathbb{R}^d$, define
the population within-cluster sum of squares as
\[
  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
\]
Define
\(
  \sA_k
    =
    \{ A \subset \mathbb{R}^d : \text{$A$ contains $k$ or fewer points} \}
\)
and set
\(
  m_k(P) = \inf_{A \in \sA_k}  W(A, P). % : \text{$A$ contains $k$ or fewer points} \}.
\)
Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and if $P_n$
is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that the
minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
(this fails, for example, with a bivariate normal distribution with identity
covariance whenever $k \geq 2$).  It is straightforward to generalize his
result to cases with multiple maximizers.  To do so, we will need the
following three lemmas.  The proofs of these lemmas are modifications of
arguments given in Sections~3 and~4 of Pollard's article.

\begin{lemma}[Continuity]\label{lem:pollard-continuous}\label{lem:pollard-last}
If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
metric $H(\cdot, \cdot)$, defined for compact subsets $A, B$ of $\mathbb{R}^d$
by $H(A,B) < \delta$ if and only if every point of $A$ is within Euclidean
distance $\delta$ of at least one point of $B$, and vice versa.
\end{lemma}
\begin{proof}
This is a slight extension of the argument given at the end of Pollard's
Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
\begin{align*}
  W(A, P) - W(B, P)
    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
\end{align*}
where $M = \int \|x\|^2 P(dx)$.  Thus,
\[
  |W(A, P) - W(B, P)|
    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
\]
Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
continuous at $A$.
\end{proof}

@@

We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
infimum over $\sE_k$.  This follows from the extreme value theorem, since
$\sE_k$ is compact and, by Lemma~\ref{lem:pollard-continuous}, the map $A
\mapsto W(A,P)$ is continuous over $\sA_k$.
\end{proof}

@@

\begin{lemma}[Uniform SLLN]\label{lem:pollard-uniform}
If $\int \|x\|^2 P(x) < \infty$, then
\(
  \sup_{A \in \sE_k} |W(A, P) - W(A,P_n)| \to 0\text{ a.s.}
\)
\end{lemma}
\begin{proof}
Pollard proves this result at the beginning of his Section~4.
\end{proof}

\begin{theorem}[$k$-Means Consistency]
Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
\end{theorem}
\begin{proof}
The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains $\sbA_k$.

The set $\sE_k$ is compact and the map $f: \sE_k \to \mathbb{R}$ with $f(A) =
W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:pollard-uniform} ensures
that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
$W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an $\bar A_n
\in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.

Now, define $\mathcal{\bar A} = \mathcal{\bar A}(k)$ to be the class of sets
$\bar A$ with $k$ or fewer points such that $W(\bar A,P) = m_k(P)$.
Lemmas~\ref{lem:pollard-compact} and~\ref{lem:pollard-continuous} together
with the extreme value theorem imply that this class is nonempty.

>>>>

Our proof relies on results from \citet{pollard1981strong}, who analyzed
$k$-means and showed that the within-cluster sum of squares converges to a
population quantity.  Specifically, for any finite set $A \subset
\mathbb{R}^d$ and probability measure $P$ supported on $\mathbb{R}^d$, define
the population within-cluster sum of squares as
\[
  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
\]
Define
\(
  \sA_k
    =
    \{ A \subset \mathbb{R}^d : \text{$A$ contains $k$ or fewer points} \}
\)
and set
\(
  m_k(P) = \inf_{A \in \sA_k}  W(A, P). % : \text{$A$ contains $k$ or fewer points} \}.
\)
Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and if $P_n$
is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
converge to a minimizer of $W(\cdot, P)$.  His main theorem assumes that the
minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
(this fails, for example, with a bivariate normal distribution with identity
covariance whenever $k \geq 2$).  It is straightforward to generalize his
result to cases with multiple maximizers.

To state the consistency theorem define
\[
  \sbA_k = \{ \bar A \subseteq \mathbb{R}^d
            : W(\bar A, P) = m_k(P)
              \text{ and $\bar A$ contains $k$ or fewer points} \}.
\]
We will later show that $\sbA_k$ is nonempty and compact whenever
$\int \|x\|^2 P(dx) < \infty$.
Let $H(\cdot, \cdot)$ denote the Hausdorff metric, defined for compact subsets
$A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every point of
$A$ is within Euclidean distance $\delta$ of at least one point of $B$, and
vice versa.  We can now state the generalized version of Pollard's $k$-means
consistency theorem.

\begin{theorem}[$k$-Means Consistency]\label{lem:kmeans-consistency}
Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
\end{theorem}

The proof of this theorem relies on a series of lemmas, which are extensions
of results given in Pollard's original article.  First, we show that the map
$A \mapsto W(A, P)$ is continuous.  Next, we show that $\sbA_k$ is nonempty
and compact.  Then, it follows that for $n$ large enough, all minimizers of
$W(\cdot, A_n)$ are contained in a compact set $\sE_k$.  The final ingredient
of the proof is Pollard's uniform strong law of large numbers, which says that
$\sup_{A \in \sE_k} |W(A,P) - W(A,P_n)| \to 0\text{ a.s.}$

\begin{lemma}[Continuity]\label{lem:kmeans-continuous}
If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
metric.
\end{lemma}
\begin{proof}
This is a slight extension of the argument given at the end of Pollard's
Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
\begin{align*}
  W(A, P) - W(B, P)
    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
\end{align*}
where $M = \int \|x\|^2 P(dx)$.  Thus,
\[
  |W(A, P) - W(B, P)|
    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
\]
Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
continuous at $A$.
\end{proof}

@@

We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
infimum over $\sE_k$.  This follows from the extreme value theorem, since
$\sE_k$ is compact and, by Lemma~\ref{lem:kmeans-continuous}, the map $A
\mapsto W(A,P)$ is continuous over $\sA_k$.
\end{proof}

@@

\begin{lemma}[Uniform SLLN]\label{lem:kmeans-uniform-slln}
If $\int \|x\|^2 P(x) < \infty$, then
\(
  \sup_{A \in \sE_k} |W(A, P) - W(A,P_n)| \to 0\text{ a.s.}
\)
\end{lemma}
\begin{proof}
Pollard proves this result at the beginning of his Section~4.
\end{proof}

With Lemmas~\ref{lem:kmeans-continuous}--\ref{lem:kmeans-uniform-slln}, we can
proceed to the proof of Theorem~\ref{lem:kmeans-consistency}.

\begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]

The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
$\sbA_k$.

The set $\sE_k$ is compact and the map $f: \sE_k \to \mathbb{R}$ with $f(A) =
W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln} ensures
that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
$W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an $\bar A_n
\in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
\end{proof}

Now, define $\mathcal{\bar A} = \mathcal{\bar A}(k)$ to be the class of sets
$\bar A$ with $k$ or fewer points such that $W(\bar A,P) = m_k(P)$.
Lemmas~\ref{lem:pollard-compact} and~\ref{lem:kmeans-continuous} together
with the extreme value theorem imply that this class is nonempty.
